Scalable Clustering of Genotype Information using MapReduce

### 2015-03-11 doc/publications/journal2014/paper.tex
Added methods, results and figure relevant to Gephi visualisation

\subsection*{Result visualisation}
% DB: in the conclusion we discussed that we provide a "toolset to interpret Mahout's clustering output". 
% DB: can we discuss here how we get a visual representation from the clustering (ClusterDumper+Gephi) and what the image tells us..

@@

@@

\begin{figure}[!tpb]%figure1
\centerline{\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{signature}}
        \label{fig:sign}
        \caption{{\bf Illustration of Mahout-based clustering of genotypes.}
      The image shows how the here introduced interface converts the VCF file to a mahout usable vector-based data type. Based on k-means it illustrates how the Map-step procedure groups of vectors by their nearest cluster center and the Reduce-step averages the vectors in each group to find the new value of the updated cluster center. The mahout-produced output is then converted into a visualisation.}

>>>>

\begin{figure}[!tpb]%figure1
\centerline{\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{gephi-p3}}
        \label{fig:sign}
        \caption{{\bf Visualisation of k-means clusters derived from the phase 3 dataset.}
        Nodes are coloured by super population: AFR (green), AMR (blue), EAS (pink), EUR (red) and SAS (cyan). 
}
\end{figure}

\subsection*{Result visualisation}
% DB: in the conclusion we discussed that we provide a "toolset to interpret Mahout's clustering output". 
% DB: can we discuss here how we get a visual representation from the clustering (ClusterDumper+Gephi) and what the image tells us..
GraphML files generated using Mahout Cluster Dumper can be visualised using popular freely-available tools such as Gephi~\cite{ICWSM09154}. An example visualisation using the phase 3 dataset, with nodes coloured by super population, is shown in Figure XX.

The clusters generated by k-means clustering and their population structure are readily apparent. Three clusters are homogeneous, containing only EAS (1 cluster) or SAS individuals (2 clusters, the second small and unconnected). Two clusters contain predominantly one super population plus a minor component of another: AMR + AFR or AFR + AMR. The sixth cluster is predominantly a mix of EUR + AMR, with 3 AFR individuals and 1 SAS.

@@

\subsection*{Graph visualisation}
GraphML files generated using Mahout Cluster Dumper were visualised using Gephi. Layouts were applied in the following order: Random, ForceAtlas 2, Yifan Hu Proportional. For ForceAtlas 2 behaviour alternatives ``dissuade hubs'', ``linlog mode'' and ``prevent overlap'' were set to true, ``approximate repulsion'' was set to false.

@@

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{signature}}
        \label{fig:sign}
        \caption{{\bf Illustration of Mahout-based clustering of genotypes.}
      The image shows how the here introduced interface converts the VCF file to a mahout usable vector-based data type. Based on k-means it illustrates how the Map-step procedure groups of vectors by their nearest cluster center and the Reduce-step averages the vectors in each group to find the new value of the updated cluster center. The mahout-produced output is then converted into a visualisation.}
